\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-10-04 chapters/gaze_independence.tex
Rewrite sota

\epigraph{%
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}
\todohsm{move introduction of paper to paper}

\todo{spatially vs temporally organized interfaces}
\todo{talk about covert shifts in spatial attention alpha}

@@

\begin{figure}
  \input{figures/gaze_independence/lis_venn.tikz.tex}
  \caption[Venn diagram of potential users of \ac{bci} assistive technology.]{%
    Venn diagram of potential users of assistive technology.
    Patients with \acf{sspi} are the general target population.
    If eye gaze-based solutions are not properly suited for an individual,
    they have \acf{sspgi}.
    Individuals with \ac{lis} have no muscular control left, or is is very limited,
    such as the ability communicate through a binary switching system.
    \Acp{bci} are especially relevant for the last two groups.
  }
  \caption{fig:gaze_independence/venn}
\end{figure}

@@

\section{State-of-the-art}
\label{sec:gaze-independence}
\todohsm{Address more systematically.}
\todo{in a paragraph describe each approach and its up and downsides}
\todo{move form copied introduction}

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory and
somatosensory\todo{cite some}
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
Furthermore, they can suffer from increased mental effort in operation and
user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

@@

\subsection{Gaze-independent visual stimulation}
This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
Non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
Alternative visual stimulation paradigms can modulate specific \ac{erp} components
that are more sensitive to stimulation in the visual periphery~\cite{Schaeff2012,Xu2022}.
Nevertheless, these methods can suffer too from reduced information
transfer rates~\cite{Chennu2013}.
Furthermore, these systems typically require users to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.
\todo{advantages disadvantages rewrite (requires gaze fixation, low itr, not tested with split attention)}

\subsection{Gaze-independent decoding}
Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
attention can be decoded separately from gaze direction.
\textcite{Aloise2012} aimed to bridge the performance gap between covert and
overt VSA decoding performance.
They compared classical linear and non-linear \ac{erp} classifiers on a covert
attention oddball \ac{erp} paradigm dataset.
The results revealed no significant performance improvement in covert VSA
decoding for any of the investigated decoders.
More recent work has made advances in decoding lateralized covert
VSA harnessing the N2pc \ac{erp} component~\cite{Thiery2016,Reichert2020b,Wang2022},
decoding covert VSA shifts from spectral content~\cite{Tonin2013}, or hybrid
stimulation paradigms~\cite{Egan2017}.
Since these methods often require a slower stimulation pace than the classical
\ac{erp} \ac{bci} paradigm, gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
leveraging spatial attention remains underexplored.

\todo{
  Complete with refs from

Fast detection of covert visuospatial
attention using hybrid N2pc and SSVEP
features
}

@@

The general goal of this PhD is to tackle the gaze-dependence in problem in the
visual oddball \ac{bci}.
\todo{refer to ITRs from previous chapter}
Section~\ref{sec:gaze-independence} showed that visual \ac{bci} paradigms form
a good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}\todo{cite},
where targets or small sets of targets are shown consecutively.
Traditionally, spatially organized interfaces have a higher information transfer
rate, but are also more gaze-dependent than temporally organized interfaces.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from the position of gaze,
which do not necessarily need to coincide.

>>>>

\epigraph{%
  ``\elide\ This begs the question: if a patient had reliable control
  over their eye gaze, why would one aim to facilitate communication with that
  patient via a BCI? It would be far simpler and more efficient to use an
  eye-tracking system.
  And even if a BCI were the preferred or only available option for some reason,
  designs requiring eye gaze control could not be used
  by patients with severe motor impairment, as these patients
  cannot direct their gaze.
  Because these are the patients who have the greatest need for BCIs, it is
  important that accurate gaze-independent BCIs exist.''
}{%
  \textcite{Egan2017}
}

\todo{talk about covert shifts in spatial attention alpha}

@@

\begin{figure}
  \input{figures/gaze_independence/lis_venn.tikz.tex}
  \caption[Venn diagram of potential users of \ac{bci} assistive technology.]{%
    Venn diagram of potential users of assistive technology.
    Patients with \acf{sspi} are the general target population.
    If eye gaze-based solutions are not properly suited for an individual,
    they have \acf{sspgi}.
    Individuals with \ac{lis} have no muscular control left, or is it is very limited,
    such as the ability communicate through a binary switching system.
    \Acp{bci} are especially relevant for the last two groups.
  }
  \label{fig:gaze_independence/venn}
\end{figure}

@@

\section{State-of-the-art}
\label{sec:gaze-independence}
\todohsm{Address more systematically.}
\todo{in a paragraph describe each approach and its up and downsides}
\todo{rewrite}

\subsection{Gaze-independent modalities}
Gaze-independent \ac{erp}-based \ac{bci}s~\cite{Riccio2012, Aloise2012} can be realized in three
ways.
Firstly, active \ac{bci} communication paradigms relying on endogenous activation from the user
do not rely on sensory stimulation.
Examples of this are imagined movement or imagined speech paradigms.
These paradigms can yield very high information transfer
rates~\cite{Willett2021,Metzger2023}, both due to their intuitiveness and the
complexity that can be captured in commands, but often only do so when paired
with invasive recording.
Non-visual reactive paradigms that use e.g. auditory and
somatosensory\todo{cite some}
stimulation do not rely on gaze redirection exist but also result in lower information transfer
rates compared to visual paradigms.
While auditory and somatosensory stimulation on its own might yield poor
\ac{itr} \textcite{Yin2016} that it can provide added value in gaze-independent
settings when coupled with oddball stimulation in a hybrid paradigm.
Their solution using the P3 components of bimodal stimulation reached an \ac{itr} of 14.94bits/min.
In general, their approach is interesting since they follow a philosophy that
relies on establishing as many communication pathways as possible.
Yet, auditory and somatosensory \acp{bci} suffer from increased mental effort in
operation and from user-dependent variability~\cite{Severens2014,Reichert2020b}.
\textcite{Severens2014} showed that the visual Hex-o-Spell~\cite{Treder2010}
outperformed a somatosensory alternative in participants with \ac{als} whose
eye motor capabilities were effectively impaired.

@@

\subsection{Gaze-independent visual stimulation}
This leads us to the second approach, current visual stimulation paradigms can be optimized, e.g. such that
the stimuli are always present in the field of view either overtly~\cite{Acqualagna2013, Won2018,
Lin2018} or covertly~\cite{,Pires2011,Lees2018}. Some noteworthy examples include the GIBS Block Speller~\parencite{Pires2011},
the GeoSpell interface~\parencite{Aloise2012}, and the \ac{rsvp}
speller~\parencite{Acqualagna2011}.
The \ac{rsvp} paradigm in particular is a prime contender for a performant gaze-independent
\ac{bci}.
\textcite{Lin2018} reported an average online \ac{itr} of 20.259 bits/min using
a paradigm with subsequently flashing sets of three characters filling the field of view of the user.
We make a distinction here between spatially organized interfaces, where
multiple targets are displayed at the same time at different spatial locations,
and temporally organized interfaces, like \ac{rsvp}\todo{cite},
where targets or small sets of targets are shown consecutively.
The Hex-o-Spell, Geospell and GIBS Block Speller are spatially organized, while
variants of the \ac{rsvp} paradigm are temporally organized.
\textcite{Lin2018} follow the philosophy that spatially organized interface
have historically shown not to perform well in the presence of gaze impairment.
However, we argue that, spatially organized interfaces in general have a much higher
\ac{itr} in healthy controls.
For instance, a regular matrix-based speller can achive an \ac{itr} to up to
30bits/min.
With the right set of adaptations, spatial attention could potentionally also
bring added value to individuals with \ac{sspgi}.
While spatial interfaces are indeed more gaze-dependent than temporal
interfaces they also provide an extra channel that can be used to transfer
information, i.e.\ the location of the stimulus.

As an alternative to spatial attention, non-spatial visual attention (feature attention) can also be exploited, such as
attention to stimulus color, shape or symbol~\cite{Zhang2010,Treder2011,Hwang2015}.
The \ac{rsvp} speller is already an example of this, relying on the user to
attend to attend the appearance of the intended character to type.
Visual stimulation paradigms relying on alternative types of attention
can modulate specific \ac{erp} components
that are more sensitive to stimulation in the visual
periphery~\cite{Schaeff2012,Xu2022}.\todo{discuss Xu2022}
Nevertheless, these methods can suffer too from reduced information
transfer rates~\cite{Chennu2013}.
Furthermore, the previously mentioned systems typically were only tested in
settings where the user was required to focus on a central fixation point while
selecting peripheral targets.
This entails that they still rely to some extent on
eye motor control, often necessitating central gaze fixation.

\subsection{Gaze-independent decoding}
Thirdly, stimuli can be presented in a standard \ac{bci} paradigm, but visuospatial
attention can be decoded separately from gaze direction.
\textcite{Aloise2012} aimed to bridge the performance gap between covert and
overt VSA decoding performance.
They compared classical linear and non-linear \ac{erp} classifiers on a covert
attention oddball \ac{erp} paradigm dataset.
The results revealed no significant performance improvement in covert VSA
decoding for any of the investigated decoders.

More recent work has made advances in decoding lateralized covert
VSA harnessing the N2pc \ac{erp}
component~\cite{Thiery2016,Reichert2020b,Wang2022}
The amplitude of the N2pc component in the contralateral hemisphere, is directly modulated
by the location of covert \ac{vsa}.
Another approach is directly decoding covert shifts in \ac{vsa} from spectral
content~\cite{Tonin2013}.
These methods are promising, but to date often require a slower stimulation pace than or cannot display as
many targets as the classical \ac{erp} \ac{bci} paradigms.
As with the alternative modality approach, they are currently only helpful as
extra communication channels in a hybrid paradigm.
\textcite{Xu2016} reached an \ac{itr} of 23.56bits/min by detecting the N2pc
component while performing \ac{ssvep} stimulation.
\textcite{Egan2017} also showed an increase in \ac{ssvep} performance by
concurrently decoding the covert \ac{vsa} shift from spectral content, but
their study only used a few stimulation targets.
In general, we can state that gaze-independent decoding in the fast-paced reactive \ac{erp} paradigm
leveraging spatial attention remains underexplored.

@@

The general goal of this PhD is to tackle the gaze-dependence in problem in the
visual oddball \ac{bci}.
\todo{refer to ITRs from previous chapter}
Section~\ref{sec:gaze-independence} showed that visual \ac{bci} paradigms form
a good candidate for a performant gaze-independent \ac{bci}.
While the central gaze fixation of most visual gaze-independent paradigms still
relies to some extent on eye motor control, we aim to circumvent this.
In such an interface, stimuli can be presented in a standard \ac{bci} paradigm, but
visuospatial attention can be decoded separately from the position of gaze,
which do not necessarily need to coincide.
